---
title: 'Meet the {GGally} package'
author: Allison Horst
date: '2020-07-07'
subtitle: ''
summary: '{GGally} for visual overviews of multivariate data'
tags:
  - coding-demo
  - study-group
  - forcats
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(GGally)
library(palmerpenguins)
library(skimr)
```

The delightful [{GGally}](https://www.rdocumentation.org/packages/GGally/versions/1.5.0) package, by [Barret Schloerke](http://schloerke.com/) and other contributors, makes it easier to explore relationships and patterns for multivariate data. In the following examples, we show several {GGally} functions to visualize your data using the [{palmerpenguins}](https://allisonhorst.github.io/palmerpenguins/) package (install from GitHub for now -- hopefully on CRAN soon). 

```{r, eval = FALSE}
library(tidyverse)
library(palmerpenguins)
library(skimr)
library(GGally)
```

## 1. Check out the penguins data! 

First, some of our go-to ways to explore data, using the **penguins** object from {palmerpenguins}. The data were originally collected and published by [Dr. Kristen Gorman](https://www.uaf.edu/cfos/people/faculty/detail/kristen-gorman.php) and the [Palmer Station LTER](https://pal.lternet.edu/). Structural size measurements for nesting adults of three penguin species (Adélie, Chinstrap and Gentoo) were recorded on islands in Palmer Archipelago, Antarctica, from 2007 - 2009. 

```{r}
# Bring it up in a new tab
# View(penguins)

# Summarize it 
summary(penguins)

# Glimpse it
tibble::glimpse(penguins)

# Skim it 
skimr::skim(penguins)
```

## 2. Visual overview of relationships with {GGally} package

More on {GGally}: https://ggobi.github.io/ggally/

Summary: "GGally extends ggplot2 by adding several functions to reduce the complexity of combining geoms with transformed data. Some of these functions include a pairwise plot matrix, a scatterplot plot matrix, a parallel coordinates plot, a survival plot, and several functions to plot networks."

### A. `ggpairs()`: a pair plot!

Out of the box: 

```{r}
ggpairs(penguins)
```

That should make you go "What." It's cool but a bit much. We can specify what we want in our pairs plot, and map color onto a variable to separate groups.

```{r}
penguins %>% 
  select(species, bill_length_mm:body_mass_g) %>% 
  ggpairs(aes(color = species))
```

### B. `ggbivariate()`: plot outcome variable with several explanatory variables

```{r}
penguins %>% 
  ggbivariate(outcome = "body_mass_g", 
              explanatory = c("species","sex", "island","flipper_length_mm"))
```

An example with custom formatting: 

```{r}
penguins %>% 
  ggbivariate(outcome = "species", 
              explanatory = c("flipper_length_mm", "island", "sex"),
              rowbar_args = list(
    colour = "purple",
    size = 4,
    fontface = "bold",
    label_format = scales::label_percent(accurary = 1)
  )) +
  scale_fill_brewer(palette = 10) +
  theme_minimal()
```

### C. `ggnostic()`: Model diagostics

```{r}
# Make a model: 
penguin_lm <- lm(body_mass_g ~ flipper_length_mm + bill_length_mm + bill_depth_mm, data = penguins)

# Look at the diagnostics! 
ggnostic(penguin_lm)
```

### D. `ggtable()`: Nicely formatted counts tables

```{r}
# Counts, for each species, tallied by island & sex
penguins %>% 
  ggtable("species", c("island","sex"))
```

And much more! Check out {GGally} for more tools to explore and visualize your multivariate data!  
